Modeling the Java Card API in B

### 2012-04-02 trunk/papers/B2012-work/javacardAPIpaper.tex
quase - arrumada a apresentacao de java card, quebrada a secao 3, incluida explicacao do modelo de Object e cia e seu uso.

The main component of the Java Card platform (Figure~\ref{fig:jcre})
is its runtime environment (JCRE), composed of a Java Card Virtual
Machine (JCVM), an API, and, usually, system and
industry-specific classes~\cite{chen:2000}. The JCRE acts as a small
operating system, being responsible for the control of the application
lifetime, security and resource management.

The Java Card API provides the means to deal with the APDU protocol,

\paragraph{Java Card applets}\label{sec:applets}

A Java Card applet is a class that inherits the
\emph{javacard.framewok.Applet} class of the Java Card API and is
implemented upon the Java Card subset of Java.  During the applet
conversion for card installation, a verification phase is performed to
check conformance of the classes to Java Card restrictions.

The current usual Java Card specification (2.2.x) allows two kinds of
applets.  The older, and most commonly used, kind of applet
manipulates directly the APDU protocol packages while the newer one
abstracts from the lower level protocol using Remote Method Invocation
(RMI). In this paper we will call the lower level applet \emph{APDU
  applet} and the higher level one, \emph{RMI applet}. RMI introduces
a layer of abstraction above the APDU protocol, and hence is usually
less efficient than APDU applets.

@@

Among the different components of the Java Card platform, the Java
Card API is the one without which no refinement into Java Card code
could be done.  Suppose, for instance, that we had the B abstract
machine of Fig.~\ref{lis:buscard}, which models a bus card with a
variable that keeps track of the current credit the user has
(\texttt{balance}) and an operation to add credits to this balance
(\texttt{add}). This simple example provides enough evidence of the
importance of having a B model of the Java Card API.  B tools can directly 
generate code for it in C, for instance, because it  is a deterministic specification and all data in
it is implementable in C.  The
resulting code would be very similar to the original B machine.

@@

However, Java Card code corresponding to the operation \texttt{add} of
this B specification would look like the one presented in
Fig.~\ref{lis:addjavacard}. In a rigorous development process, we do
not want the code generator to carry out all these non-verified
transformations in the code. So, we need to be able to refine the
original B machine into something such as the B refinement of
Fig.~\ref{lis:addimplementation}. Then, code generation would be
straightforward (as it should be).

@@

We can identify two main kinds of transformation that have to be
carried out in order to obtain the code of operation \texttt{add} in
figs.~\ref{lis:addjavacard} and~\ref{lis:addimplementation} from the
abstract specification in fig.~\ref{lis:buscard}:

@@

The Java Card API, in its version 2.2.2\footnote{Java Card 2.2.2,
  although not the most recent version of the Java Card platform, is
  still widely used, and can easily be extended to Java Card 3 Classic
  Edition, a backward compatible evolution of version 2.2.2.,
  including some bug fixes, clarifications against the Java Card
  v2.2.2 specifications and new security algorithms (Java Card 3
  Platform Release Notes, available at
  http://www.oracle.com/technetwork/java/javame/javacard/download/overview/index.html).}, is composed of 14
packages, containing 93 classes and interfaces, as shown in
Table~\ref{tab:packages}. Their interfaces have all been modeled in B,
and the resulting models are available at the KitSmart project
page\footnote{https://code.google.com/p/kitsmart/}.  Differently from
JML and OCL, though, B is not object oriented, and some modeling
solutions had to be found to cope with this lack of resources.  We
present in the following the main characteristics and some
restrictions of our models:

@@

\item The hierarchy of the Java Card API objects is partially modeled
  by abstract set inclusion. For instance, the class Throwable extends
  the class Object, in Java Card. In our model, we  have that the
  machine Throwable SEES the machine Object, and too show that all throwables are indeed
  objects, a property stating that THROWABLE is a subset of OBJECT is
  included, associated to the Throwable machine.

@@

As previously stated, we chose to model exception raising conditions
as operation pre-conditions in the B models. For instance, if the Java
Card specification states that a method should throw an exception in
case a parameter array is empty, the pre-condition of the
corresponding B operation will require that this parameter is not
empty. The B method then requires that every call to that operation
satisfies the operation's pre-condition. As a consequence, code that
is formally generated using the B method and the Java Card API
specification should never raise this specific exception, as it is
statically proved that all calls to this operation are made with
non-empty parameter arrays. Not all situations leading to exceptional
conditions are modeled, however, as some of them depend on the Java
Card Runtime Environment (JCRE) and its state, which has not yet been
modeled, and some require a more detailed specification of the
contents of some data and other internal behavior. In these cases, the
exception raising conditions are only left as comments to the B
specification, and can either be used as additional documentation or be
considered in future work.

@@

MACHINE ThrowableSet
SEES 
    ObjectSet    
CONSTANTS  THROWABLE
PROPERTIES THROWABLE <: OBJECT
END
\end{lstlisting}

>>>>

The main component of the Java Card platform
is its runtime environment (JCRE), composed of a Java Card Virtual
Machine (JCVM), an Application Programming Interface (API), and, usually, system and
industry-specific classes~\cite{chen:2000}. The JCRE acts as a small
operating system, being responsible for the control of the application
lifetime, security and resource management.

The Java Card API is the part of the Java Card platform that a developer of any Java Card application must use. It defines the constituents of an applet, provides the means to deal with the APDU protocol and communicate data and commands between the host application and the card, and defines some basic services for security and cryptography implementation.

In a Java Card application, the applet is the part which defines the card provided services. 
An applet inherits the \emph{javacard.framewok.Applet} class of the Java Card API and is
implemented upon the Java Card subset of Java to provide these services.  During the applet
conversion for card installation, a verification phase is performed to
check conformance of the classes to Java Card restrictions.

@@

\paragraph{Motivation}
Among the different components of the Java Card platform, the Java
Card API is the one without which no refinement into Java Card code
could be done.  Suppose, for instance, that we had the B abstract
machine of Listing~\ref{lis:buscard}, which models a bus card with a
variable that keeps track of the current credit the user has
(\texttt{balance}) and an operation to add credits to this balance
(\texttt{add}). This simple example provides enough evidence of the
importance of having a B model of the Java Card API.  B tools can directly 
generate code for it in C, for instance, because it  is a deterministic specification and all data in
it is implementable in C.  The
resulting code would be very similar to the original B machine.

@@

However, Java Card code corresponding to the operation \texttt{add} of
this B specification would look like the one presented in
Listing~\ref{lis:addjavacard}. In a rigorous development process, we do
not want the code generator to carry out all these non-verified
transformations in the code. So, we need to be able to refine the
original B machine into something such as the B refinement of
Listing~\ref{lis:addimplementation}. Then, code generation would be
straightforward (as it should be).

@@

We can identify two main kinds of transformation that have to be
carried out in order to obtain the code of operation \texttt{add} in
figs.~\ref{lis:addjavacard} and~\ref{lis:addimplementation} from the
abstract specification in Listing~\ref{lis:buscard}:

@@

\paragraph{The model}
The Java Card API, in its version 2.2.2\footnote{Java Card 2.2.2,
  although not the most recent version of the Java Card platform, is
  still widely used, and can easily be extended to Java Card 3 Classic
  Edition, a backward compatible evolution of version 2.2.2.,
  including some bug fixes, clarifications against the Java Card
  v2.2.2 specifications and new security algorithms (Java Card 3
  Platform Release Notes, available at
  http://www.oracle.com/technetwork/java/javame/javacard/download/overview/index.html).}, is composed of 14
packages, containing 93 classes and interfaces, as shown in
Table~\ref{tab:packages}. Their interfaces have all been modeled in B,
and the resulting models are available at the KitSmart project
page\footnote{https://code.google.com/p/kitsmart/}.  Differently from
JML and OCL, though, B is not object oriented, and some modeling
solutions had to be found to cope with this lack of resources.  We
present in the following the main characteristics and some
restrictions of our models:

@@

\item The hierarchy of the Java Card API objects is partially modeled
  by abstract set inclusion. For instance, the class Throwable extends
  the class Object, in Java Card. In our model, we  have that the
  machine Throwable SEES the machine Object, and to show that all throwables are indeed
  objects, a property stating that THROWABLE is a subset of OBJECT is
  included, associated to the Throwable machine.

@@

A small piece of specification with some of the characteristics described above is presented in Listing \ref{lis:object}. It presents two \texttt{java.lang} classes: \texttt{Object} and \texttt{Throwable}.  \texttt{Object} is  superclass to almost all API classes, with its only method  \texttt{equals} that checks if an object (from which a reference is received as argument) is the same as the current one. \texttt{Throwable} extends \texttt{Object}, but does not define any other method nor redefines \texttt{equals}. The inheritance hierarchy is partially simulated in B by the corresponding set inclusion. These abstract sets are then used to refer to the corresponding class as a type as can be seen in Listing \ref{lis:class-types}.

@@

MACHINE ThrowableSet
SEES 
    ObjectSet    
CONSTANTS  THROWABLE
PROPERTIES THROWABLE <: OBJECT
END

\end{lstlisting}

\lstset{language=B, caption={Using classes as types},label=lis:class-types}
\begin{lstlisting}
op(anObject,otherObject) = 
	PRE	
		anObject : OBJECT & otherObject : THROWABLE
...
\end{lstlisting}

\paragraph{Discussions on the model}
As previously stated, we chose to model exception raising conditions
as operation pre-conditions in the B models. For instance, if the Java
Card specification states that a method should throw an exception in
case a parameter array is empty, the pre-condition of the
corresponding B operation will require that this parameter is not
empty. The B method then requires that every call to that operation
satisfies the operation's pre-condition. As a consequence, code that
is formally generated using the B method and the Java Card API
specification should never raise this specific exception, as it is
statically proved that all calls to this operation are made with
non-empty parameter arrays. Not all situations leading to exceptional
conditions are modeled, however, as some of them depend on the Java
Card Runtime Environment (JCRE) and its state, which has not yet been
modeled, and some require a more detailed specification of the
contents of some data and other internal behavior. In these cases, the
exception raising conditions are only left as comments to the B
specification, and can either be used as additional documentation or be
considered in future work.
